Challenges and Solutions of Developing and Implementing a Novel Desktop-as-a-Service

### 2024-05-03 paper.tex
Some language improvements

\subsubsection{Network Latency and Bandwidth}
Network Latency and Bandwidth are crucial factors
in delivering web-based access to DaaS apps,
directly impacting system responsiveness experienced by users.
Supporting standardized frame-buffer protocols enables user interaction initially
and provide additional access to external devices like printers or USB sticks.
In cases involving for example video content transmission,
network bandwidth often becomes the main bottleneck,
especially with high-resolution video streams.
In contrast to that,  when transmitting audio, mouse movements,
or keyboard inputs, network latency becomes more important.

Application-specific requirements can be managed to some extent
through protocol-inherent tuning parameters and compression algorithms.
Our solution therefore offers a range of tuning options
in order to adapt to different use cases
and empowering users to choose their own tailored solution.

We utilize the Guacamole desktop gateway implementation,
supporting widely-used frame-buffer protocols like VNC and RDP,
along with relevant tuning options.
Guacamole also enables audio support and integration with external devices
through a JavaScript-based web client (guacamole.js).
As depicted in figure~\ref{websocket_proxy},
the proxy component recives HTTP requests from the guacamole.js webclient
and forwarded by the backend by using regular WebSocket opcodes (Guacamole protocol).
Such communication is forwarded to the Guacamole daemon (guacd)
which in turn communicates to each instance
by using the frame-buffer protocol of choice (VNC or RDP).

However, integrating Guacamole into the system
requires consideration of additional factors besides the frame-buffer protocols itself.
Authentication mechanisms for example are particularly important in a DaaS context,
as users expect seamless authentication across the whole system
without multiple logins per session.
Additionally, safeguarding sensitive information like user passwords is crucial
for maintaining system security and user trust.
To mitigate the risk of unintended system manipulation, Guacamole was integrated
in a proxied approach and by using websockets.
This ensures that credentials of a generated instance remain secret to the host system
while still providing authorized, passwordless access to the hosted application.

Moreover, the proxy mechanism enables support for other additional features
such as dynamic desktop resizing.
While Guacamole offers support for that by default in RDP,
support is completely missing for the VNC protocol.
Nevertheless, the proxied approach enables to intercept all exchanged websocket opcodes
and by listening for size requests
the requested feature could be added
by directly interacting with the instance on another layer.

@@

\subsubsection{Internal processing and Communication}
Another critical aspect directly related to performance and responsiveness
is how data is internally processed, stored, and communicated
among all involved components and instances.
This includes metadata collected and processed by essential components
and involves direct exchange of raw data with object instances.

Ensuring a satisfying user experience in a DaaS system
implies specific components in a host system.
This includes access to all backend services
used to create containers and virtual machines,
local or remote databases, authentication services, filesystems, and network devices.
Additionally, a web server is integrated to provide relevant API endpoints
for frontend communication and websocket endpoints for viewer communication.

Generally spoken, all of these components can potentially become bottlenecks
when handling high throughputs or large amounts of concurrent applications
requiring low latency access and especially if all necessary backend components
are served from the same hardware.
Architecturally, all relevant components
must therefore be designed for distribution across different hosting systems.
However, retaining certain components directly on the host system
may still be required to ensure low-latency access.

From our observations, components requiring realtime and low-latency performance,
should physically be kept as close to the core hosting system as possible
as shown in figure~\ref{perf_internal_communication}.
This includes all data and communication relevant to
audio, video, mouse, and keyboard transmission,
as well as all backend services for hosting containers and virtual machines.
Ideally, these components should be hosted natively on the host system
for virtual network and hardware access
or at least be connected to the same physical network
in order to minimize network latencies.

Regarding other components such as database access,
authentication providers, and filesystems,
maintaining a close distance from the host is advisable,
but this depends largely on the total number of users
as well as their specific use cases.
On smaller scales, keeping these components
in close physical proximity to the host remains prudent.

@@

To comply with both approaches,
we designed the system and its backend services
to be split into logical units that can be distributed across dedicated systems.
Each backend service receives dedicated configuration
and can be accessed via hostname and port tuples.
By default, all backend services reside on the same host,
with tuples derived from the host configuration.
In a distributed context, services can be distributed to dedicated remote systems
to adhere to horizontal scaling principles.
Relevant tuples can then be derived from any node
using a common database containing meta information.

Vertical scaling of DaaS nodes is not currently planned,
but administrators are free to use custom third-party toolchains,
such as GPU virtualization, for vertical scaling on a per-host basis.
Although vertical scaling within a host
is beyond the scope of our research project,
scaling the instances hosted within the system is feasible for both dimensions.
For vertical scaling of instances, we provide appropriate parameters
and API endpoints to manipulate resource assignments,
allowing for arbitrary configurations.

For horizontal scaling of instances, the built-in mechanisms
of each backend service are automatically configured
and utilized to distribute instances across a pool of connected nodes.
For virtual machines, Proxmox implements such mechanisms
as part of its `Proxmox VE High Availability (HA)' cluster.
For containers, similar concepts can be integrated with any backend service,
such as Kubernetes.
Within these clusters,
DaaS instances are entirely agnostic to their hosting system
and only need to be reachable
via a known hostname or IP as a single requirement.
As shown in figure~\ref{scale_objects_horizontal},
we provide a globally maintained state of data, authentication and storage
which is accessible either by horizontally scaled backends and frontends
or otherwise,
by horizontally and vertically scaled instances in our DaaS object domain.

@@

	      by distributing requests and responses to their respecitive recipients

@@

	      and is therefore independent of surrounding components and contexts.

@@

	      and remains fully agnostic of surrounding components

@@

	      a customizable amount of redundant systems may be configured.

@@

	      enables migration and redundancy of instances or in other words, horizontal scaling.

@@

If that is the case, each distributed DaaS node might only contain essential and time-critical components,
including the backend service hosting relevant instances as well as the described
websocket proxy connected to the guacamole daemon and the user.
Any other component may reside elsewhere within the DaaS network.

@@

For Linux operating system deployments,
the free software projects xrdp and x11vnc enable all desired features
and single-app functionality can be easily obtained
by using a custom window manager such as for example qtile
~\footnote{\url{https://qtile.org/}}.

@@

One challenge was imposed by Guacamole's current lack of support
for dynamic desktop resizing with VNC,
leading to misaligned layouts and a poor user experience.
To address this until such features are available in Guacamole
\footnote{\url{https://issues.apache.org/jira/browse/GUACAMOLE-1196}},
we implemented a separate mechanism using a websocket proxy
and direct interaction with each instance.

@@

Another challenge was managing RDP sessions which are typically tied to actual users,
unlike VNC sessions which are typically authenticated against
a server-maintained user registry.
Utilizing the websocket proxy, the backend could pass credentials
to the server without revealing them to the user.

Essentially, Guacamole with its protocol implementations and additional features like audio and printing,
provided the fundamental base for our core capabilities.

Finally mentioned, certain instances might not be fully accessible during runtime
due to hardware malfunctions, inappropriate system configurations for simply during system updates.
Therefore, as a failover mechanism additional logic had to be added
in order to provide reasonable debug views and recovery mechanisms,
e.g.~to recreate instances from a specific snapshot or configuration state.

A crucial lesson learned was therefore to realize that
while individual problems can mostly be solved
using established paradigms, services or libraries with only slight modifications,
the real challenge for a full-fledged DaaS system lies in it's abilities
to combining such isolated solutions appropriately
and for each specific infrastructure and platform configuration.

@@

This work was funded by the Federal Ministry for Economic Affairs and Climate Action
('\textsl{Bundesministerium f\"ur Wirtschaft und Klimaschutz}')
in the framework of the central innovation programme
for small and medium-sized enterprises
('\textsl{Zentrales Innovationsprogramm Mittelstand}').

>>>>

\subsubsection{Network Latency and Bandwidth}
Network Latency and Bandwidth are crucial factors
in delivering web-based access to DaaS apps,
directly impacting system responsiveness experienced by users.
Supporting standardized frame-buffer protocols enables user interaction initially
and provide additional access to external devices like printers or USB sticks.
In cases involving, for example, video content transmission,
network bandwidth often becomes the main bottleneck,
especially with high-resolution video streams.
In contrast to that, when transmitting audio, mouse movements,
or keyboard inputs, network latency becomes more important.

Application-specific requirements can be managed to some extent
through protocol-inherent tuning parameters and compression algorithms.
Our solution, therefore, offers a range of tuning options
to adapt to different use cases
and empowering users to choose their own tailored solutions.

We utilize the Guacamole desktop gateway implementation,
supporting widely used frame-buffer protocols like VNC and RDP,
along with relevant tuning options.
Guacamole also enables audio support and integration with external devices
through a JavaScript-based web client (guacamole.js).
As depicted in figure~\ref{websocket_proxy},
the proxy component receives HTTP requests from the guacamole.js web client
and is forwarded by the backend by using regular WebSocket opcodes (Guacamole protocol).
Such communication is forwarded to the Guacamole daemon (guacd)
which in turn communicates to each instance using the frame-buffer protocol of choice (VNC or RDP).

However, integrating Guacamole into the system
requires consideration of additional factors besides the frame-buffer protocols itself.
Authentication mechanisms, for example, are essential in a DaaS context,
as users expect seamless authentication across the whole system
without multiple logins per session.
Additionally, safeguarding sensitive information like user passwords is crucial
for maintaining system security and user trust.
To mitigate the risk of unintended system manipulation, Guacamole was integrated
using a proxied approach and WebSockets.
This ensures that the credentials of a generated instance remain secret to the host system
while still providing authorized, passwordless access to the hosted application.

Moreover, the proxy mechanism enables support for other additional features
such as dynamic desktop resizing.
While Guacamole offers support for that by default in RDP, support for the VNC protocol is completely missing.
Nevertheless, the proxied approach enables intercepting all exchanged WebSocket opcodes
and by listening for size requests
the requested feature could be added
by directly interacting with the instance on another layer.

@@

\subsubsection{Internal processing and Communication}
Another critical aspect directly related to performance and responsiveness
is how data is internally processed, stored, and communicated
among all involved components and instances.
This includes metadata collected and processed by essential components
and involves the direct exchange of raw data with object instances.

Ensuring a satisfying user experience in a DaaS system
implies specific components in a host system.
This includes access to all backend services
used to create containers and virtual machines,
local or remote databases, authentication services, filesystems, and network devices.
Additionally, a web server is integrated to provide relevant API endpoints
for frontend communication and WebSocket endpoints for viewer communication.

Generally spoken, all of these components can potentially become bottlenecks
when handling high throughputs or large amounts of concurrent applications
requiring low latency access, especially if all necessary backend components
are served from the same hardware.
Architecturally, all relevant components
must be designed for distribution across different hosting systems.
However, retaining specific components directly on the host system
may still be required to ensure low-latency access.

From our observations, components requiring real-time and low-latency performance,
should physically be kept as close to the core hosting system as possible
as shown in figure~\ref{perf_internal_communication}.
This includes all data and communication relevant to
audio, video, mouse, and keyboard transmission,
as well as all backend services for hosting containers and virtual machines.
Ideally, these components should be hosted natively on the host system
for virtual network and hardware access
or at least be connected to the same physical network
to minimize network latencies.

Regarding other components, such as database access,
authentication providers, and filesystems,
maintaining a close distance from the host is advisable,
but this depends mainly on the total number of users
as well as their specific use cases.
On smaller scales, keeping these components
in close physical proximity to the host remains prudent.

@@

To comply with both approaches,
we designed the system and its backend services
to be split into logical units that can be distributed across dedicated systems.
Each backend service receives a dedicated configuration
and can be accessed via hostname and port tuples.
By default, all backend services reside on the same host,
with tuples derived from the host configuration.
In a distributed context, services can be distributed to dedicated remote systems
to adhere to horizontal scaling principles.
Relevant tuples can then be derived from any node
using a common database containing meta information.

Vertical scaling of DaaS nodes is not currently planned,
but administrators are free to use custom third-party toolchains,
such as GPU virtualization for vertical scaling on a per-host basis.
Although vertical scaling within a host
is beyond the scope of our research project,
scaling the instances hosted within the system is feasible for both dimensions.
For vertical scaling of instances, we provide appropriate parameters
and API endpoints to manipulate resource assignments,
allowing for arbitrary configurations.

For horizontal scaling of instances, the built-in mechanisms
of each backend service are automatically configured
and utilized to distribute instances across a pool of connected nodes.
Proxmox implements such mechanisms
for virtual machines as part of its `Proxmox VE High Availability (HA)' cluster.
Similar concepts can be integrated for containers with any backend service,
such as Kubernetes.
Within these clusters,
DaaS instances are entirely agnostic to their hosting system
and only need to be reachable
via a known hostname or IP as a single requirement.
As shown in figure~\ref{scale_objects_horizontal},
we provide a globally maintained state of data, authentication, and storage
which is accessible either by horizontally scaled backends and frontends
or otherwise,
by horizontally and vertically scaled instances in our DaaS object domain.

@@

	      by distributing requests and responses to their respective recipients

@@

	      and is, therefore, independent of surrounding components and contexts.

@@

	      and remains entirely agnostic of surrounding components

@@

	      a customizable number of redundant systems may be configured.

@@

	      enables migration and redundancy of instances or, in other words, horizontal scaling.

@@

If that is the case, each distributed DaaS node might only contain essential and time-critical components,
including the backend service hosting relevant instances as well as the described
WebSocket proxy connected to the guacamole daemon and the user.
Any other component may reside elsewhere within the DaaS network.

@@

For Linux operating system deployments,
the free software projects xrdp and x11vnc enable all desired features
and single-app functionality can be easily obtained
by using a custom window manager such, as for example, qtile
~\footnote{\url{https://qtile.org/}}.

@@

One challenge was imposed by Guacamole's current lack of support
for dynamic desktop resizing with VNC,
leading to misaligned layouts and a poor user experience.
To address this until such features are available in Guacamole
\footnote{\url{https://issues.apache.org/jira/browse/GUACAMOLE-1196}},
we implemented a separate mechanism using a WebSocket proxy
and direct interaction with each instance.

@@

Another challenge was managing RDP sessions, which are typically tied to actual users,
unlike VNC sessions, which are typically authenticated against
a server-maintained user registry.
Utilizing the WebSocket proxy, the backend could pass credentials
to the server without revealing them to the user.

Essentially, Guacamole, with its protocol implementations and additional features like audio and printing,
provided the fundamental base for our core capabilities.

Finally, certain instances might not be fully accessible during runtime
due to hardware malfunctions and inappropriate system configurations during system updates.
Therefore, as a failover mechanism, additional logic had to be added
to provide reasonable debug views and recovery mechanisms,
e.g., recreate instances from a specific snapshot or configuration state.

A crucial lesson learned was, therefore, to realize that
while individual problems can mostly be solved
using established paradigms, services, or libraries with only slight modifications,
the real challenge for a full-fledged DaaS system lies in its abilities
to combine such isolated solutions appropriately
and for each specific infrastructure and platform configuration.

@@

The Federal Ministry for Economic Affairs and Climate Action funded this work
('\textsl{Bundesministerium f\"ur Wirtschaft und Klimaschutz}')
in the framework of the central innovation programme
for small and medium-sized enterprises
('\textsl{Zentrales Innovationsprogramm Mittelstand}').
